Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-09-15 manuscript/manuscript.tex
added supp section
went over first two sections

\jy{How about putting all these sections into one section on case
  study? They can be subsections in the merged section.
  After that, we can have a section on recommendations for open
  data curation based on what we learn from the case study. For each
  recommended principle, we can link to the specifics of the case
  study.}

The paper is organized as follows:
Section~\ref{sec:data} `\textit{NYC 31 Service Request Data} offers a 
look at the history of the 311system. Long-term trends are 
presented in a 10-year data analysis (2014-2023).
Section~\ref{sec:issues} \textit{Data Cleansing Issues} offers a 
discussion of specific data cleansing issues impacting data quality and 
curation efficiency. These issues include structural issues, compliance 
with the Data Dictionary, and missing, blank, or N/A entries. We also
explore field compliance with a domain of legal or acceptable values.
Next we deal with several important issues surrounding logical inconsistencies 
and concerning patterns found in the data. We also explore an issue of precision 
versus accuracy. This section concludes by identifying duplicate/redundant data fields
and Data Dictionary observation. Section~\ref{sec:recommendations} 
\textit{Recommendations} provides actionable suggestions for 
mitigating or resolving the identified issues. Section~\ref{sec:discussion} ``Discussions'' 
concludes with an encapsulation of key insights and implications of our research.

\section{NYC 311 Service Request Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system.

@@

To begin, we looked at a 10-year period (2014-2023) to
observe a longer time-frame; see
Figure~\ref{fig:10-yr-monthly}. During that time-frame,
the number of 311 SRs grew by nearly 50\%.

@@

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=0.8\textwidth]{10-year-trend_yearly.pdf}
 	\caption{10-year (2014-2023) Monthly SR Counts}
  	\label{fig:10-yr-monthly}
\end{figure}

Identifying the responsible agency is essential 
for troubleshooting discrepancies. This identification also assists in 
determining whether an issue is specific to an Agency or indicative 
of a broader, systemic problem. Six agencies account for 90\% of 
the complaints: New York Police Department (NYPD), 
Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).

@@

Figure~\ref{fig:SR_complaints} shows that the top 20 complaint 
types account for 70\% of the total 220 different types. Notably, 
these top complaints include several variations of noise-related 
issues, which collectively represent 22\% of all complaints, the 
most frequently occurring type of complaint.

@@

@@

>>>>

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over the 10-year period from 2014 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section 
also investigates field compliance with legal or acceptable values, 
highlights logical inconsistencies, and examines concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the Data Dictionary. Section~\ref{sec:recommendations} offers 
practical recommendations for mitigating or resolving these issues, 
while Section~\ref{sec:discussion} encapsulates key insights and 
discusses the broader implications of our findings.

\section{NYC 311 Service Request Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
\jy{should we mention the system upgrade just before the pandemic to
  document this historical event?}
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system.

@@

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=0.8\textwidth]{10-year-trend_yearly.pdf}
 	\caption{10-year (2014-2023) annual SR Counts.}
  	\label{fig:10-yr}
\end{figure}

Our investigation focused on a 10-year period (2014--2023).
Figure~\ref{fig:10-yr} shows the annual counts of NYC 311 SR over the
10-year period from 2014 to 2023, revealing a clear upward trend in
public engagement with the system. The most significant increase
occurred in 2020, when the number of requests surged to 3.23 million
due to the COVID-19 pandemic, illustrating the essential role of the
311 system in supporting residents during times of crisis. Overall, SR
counts increased by nearly 49.5\% between 2014 and 2023, reflecting
growing public reliance on the system for addressing non-emergency
issues. The steady rise in SRs can be attributed to the increased
accessibility of the 311 system via online and mobile platforms, as
well as heightened public awareness of the service. While the spike in
2020 was exceptional, the broader trend indicates sustained growth in
service request volume over the decade, highlighting the system's
evolving role in managing both routine city operations and
extraordinary events.

@@

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
agency, showing the cumulative percentage of SRs handled by each
agency over a given time period. This figure highlights that the
distribution of SRs is concentrated among a few key agencies. The six
largest agencies are New York Police Department (NYPD), 
Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).
They collectively handle over 90\% of the total SR volume. This is
indicative of the critical roles these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its central role in
addressing complaints related to public safety and order. The
remaining 10\% distributed across a large number of smaller
agencies. This concentration of requests points to the systemic
reliance on a few key departments for resolving most non-emergency
issues, further emphasizing the importance of optimizing data handling
and curation processes within these high-volume agencies to ensure
efficiency and responsiveness.

@@

@@

Figure~\ref{fig:SR_complaints}
illustrates the top 20 complaint types, accounting for 70\% 
of all Service Requests (SRs), and their cumulative percentage. The 
distribution of complaints is skewed, with a small number of issues 
dominating the total SR volume. Noise-related complaints, including 
residential, commercial, and street noise, make up 22\% of all 
complaints, making them the most frequent issue in the NYC 311 system. 
Other prominent categories include illegal parking, heat/hot water 
complaints, and blocked driveways. The cumulative percentage curve 
shows that after the top 20, the remaining complaint types are spread 
thinly across the remaining 30\% of SRs. This concentration suggests 
that improving responses to the most frequent complaints could have 
a significant impact on overall service efficiency and resident 
satisfaction. It also provides insights into the operational 
pressures faced by the city agencies responsible for handling these 
high-volume complaints.

@@

\section*{Supplementary Material}
The NYC 311 SR data and code used to produce the results reportred in
this paper are available [FixMe].
\jy{need to think: we need to keep a copy of the downloaded data;
  if the data portal changes and our code may not work.}
